---
title: "Appendix"
---

## Assignment Algorithm

-   The number of school slots at each school is predetermined based on capacity constraints.
-   Students are tentatively placed in the applicant pools for their first-choice schools and are ranked in descending order by SEA score within each application pool.
-   The school at which the n~j~th ranked applicant has the highest SEA score is determined to be the most highly subscribed/ranked school, this score becomes the cutoff score for this school, and the top n~j1~ students in the applicant pool for top-ranked school j1 are admitted to school j2.
-   The top-ranked school's slots and the admitted students are removed from the process, and the second choice becomes the new first choice for students who had the top-ranked school as their first choice but did not gain admission.
-   The process repeats in another round to assign students to the second highest ranked school j2 and determine the cutoff score for the second-ranked school.
-   The process repeats in subsequent rounds until all slots are filled.

## Omitted covariates

### Community
I chose to omit community as a control. Initially, I envisioned community as a confounder:
```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(igraph)

edges <- c("Community", "Wealth", "Community", "Prestige School", "Wealth", "Prestige School")
dag <- graph(edges, directed = TRUE)

# coordinates
layout <- matrix(c(0, 1, -1, 0, 1, 0), ncol = 2, byrow = TRUE)

# plot
plot(dag, layout = layout, edge.arrow.size = 0.5, vertex.size = 15)
```
However, I think Community is a mediator, meaning that it is a part of the causal pathway between wealth and increased likelihood of being assigned to a prestige school. Wealth is more likely to effect the community one lives in as opposed to the reverse. The relationship is more like the following DAG:

```{r, echo=FALSE, message=FALSE}
edges <- c("Wealth", "Community", "Community", "Prestige School")
dag <- graph(edges, directed = TRUE)

layout <- matrix(c(0, 0, 1, 1, 2, 0), ncol = 2, byrow = TRUE)

plot(dag, layout = layout, edge.arrow.size = 0.5, vertex.size = 15)
```

