---
title: "Project Proposal"
author: "Luke Gooding"

date: last-modified
institute: "University of Pennsylvania"
toc: true

link-citations: TRUE
bibliography: bib.bib

format:
  html:
    self-contained: TRUE
    code-fold: TRUE

editor: source
---

## Introduction

Trinidad and Tobago earned its independence in 1962. In that transition period from colony to independent state, education was used as a means of post-colonial nation building [@london_educational_1997]. The emerging education system, therefore, emphasized access and quality with the message being that education is central to the economic development of any nation [@london_educational_1997]. The country consequently has a robust and ever expanding public schooling system which, as of 2019, educates 92% of primary school students across 558 institutions and 88% of secondary school students across 199 institutions [@noauthor_education_2021].

Importantly, one of the major stakeholders in Trinidad and Tobago's education system are churches. Before independence, most of the nation's schools were run by various religious boards. Churches owned and managed these schools with autonomy and were allowed to charge enrolled students. With the goal of universal secondary education, the government took over the responsibility of funding these schools, which are now known as government 'assisted' or denominational schools [@london_educational_1997]. They standardized syllabi and exams across all the nation's public schools while allowing these religious boards to maintain administrative responsibility and their religious ethos. This arrangement is codified in The Concordat of 1960 [@noauthor_policy_nodate].

Having a publicly funded school system requires the government to assign graduating Primary School students to Secondary Schools. I refer to this process of assignment as the transition process. First, students choose 4 secondary schools they want to attend and rank them. Then, they do the Secondary Entrance Assessment (SEA). Based on their performance on the exam, they are placed into one of their 4 choices using an assignment algorithm (Appendix). If they perform poorly, they will be assigned to a school near to where they live.

However, according to the Concordat, denominational secondary schools have a special privilege:

*The Principals of Assisted Secondary Schools will make available a minimum of 80 per centum of the First Form entry places to those who, by passing the test, qualify on the results of the Common Entrance Examination for free secondary education...The Principals will be free to allocate up to 20 per centum, the remaining places as they see fit provided normally that the pass list of the Common Entrance Examination serves to provide the pupils. [@noauthor_policy_nodate]*

This part of the Concordat, colloquially known as the '20% list' is the subject of my research. Campbell refers to the Concordat as one of many colonial holdovers present in our education system [@london_educational_1997] that perpetuates elitism. Coloquially, we refer to elite secondary schools as 'prestige schools'.

'Prestige schools' are secondary schools that are considered high performing and are usually ranked first in students' 4 choices. Some refer to these schools as 'first choice schools'. To operationalize this term, I have devised a definition:

*If the number of students that chose a school as their first choice exceeds the capacity of that school, this school will be classified as a 'prestige school'.*

I can then order schools based on cutoff scores (similar to the way the Assignment Algorithm works). For example, if 250 students choose school X but school X has a capacity of 100, then X will be considered a prestige school with a cutoff score equal to the score of the 100th best performing student. I believe my definition is suitable because I think 'prestige' as a construct is more about perception than reality. It is possible to develop a ranking system for secondary schools based on historical academic and extracurricular performance but for the purpose of my analysis, the aggregated perception of students and their parents is more meaningful, and this is proxied by their choices. These choices set up a system of demand and supply where some schools have a greater demand than others, creating 'prestige'. It is not my goal to explain the underlying reason for this 'prestige' but rather to identify it and investigate a particular externality.

These schools also tend to be government assisted schools with the 20% list privilege. It then begs the question how this 20% list is used today. Remember, these churches went from having control over 100% of their incoming students in a time when secondary education was expensive but in low demand to now only controlling 20% of their incoming students in a time when secondary education is free but in high demand.

Unfortunately, there are not many studies identifying and focusing on 'prestige schools' in particular. C. Karibo Jackson, looked at a special subset of 'selective schools'. He showed that attending schools with higher-achieving peers increases academic performance and that short-run increases in peer achievement improve outcomes only at high-achieving selective schools[@jackson_can_2013].

A study conducted by the Inter-American Development Bank shows that there are no meaningful differences between the academic outcomes of government versus government assisted secondary schools between the ages of 10 and 15 in terms of dropout rates or examination performance at age 15 in Trinidad and Tobago [@beuermann_privately_2015]. This is despite the study also noting large differences in teacher and peer quality between school types. The research made no mention of 'prestige schools' which I theorize is an outlying subset of government assisted schools.

### Theory and Hypothesis

The existence of 'prestige schools' has many effects on individuals navigating the transition process. Failure to get into a 'prestige school' can have negative emotional, social and psychological impact on students and hence the effort to avoid this failure would cause students to experience stress, anxiety and fear [@wreo31474]. Parents also often resort to investing in supplementary tutoring in preparation for the SEA exam [@barrow_supplementary_2012].

With such high stakes, the 20% list is a powerful tool. While its purpose is to allow principals to hand pick students who share the same religion and values as their institution, many have claimed that the 20% list is a tool to allow students of wealthy families to gain admittance to 'prestige schools'. I am interested in whether or not the 20% list benefits wealthier families and generates inequality.

I hypothesize that there is a significant relationship between the existence of the 20% list and rate of admittance of students from wealthy families to 'prestige' secondary schools. More specifically, I think that attending a private primary school increases your likelihood of being admitted to a prestige school.

Given that I do not have granular data of students' family income, I think private primary school attendance is a suitable proxy for family wealth because only 8% of students belong to families that are willing to incur the cost of private schools.

### Empirical Extension

I will also test the strength of the relationship in question at different levels modelled after Jackson's research [@jackson_can_2013]. He uses a different ordering system to rank schools by mean incoming SEA scores. He then makes 3 categories: 1 to 30, 30 through 90 and Above 90 and compares the effect at each level. While my ordering system uses cutoff scores, I can model his categorization to determine if there are different strengths of the relationship between the 20% list and the rate of private school student admissions at the 1 to 10, 11-20 and 21-30 levels.

Ranking schools based on cutoff scores is a useful index for competitiveness of a school for two reasons. Firstly, in order for a school to have a cutoff score, there must be more students who choose the school as their first choice than there are spots available. Secondly, higher cutoff scores means higher performing, more competitive students.

These regressions will aid in addressing heterogeneity in my research question. It is entirely plausible that attending a private primary school has varying effects depending on how competitive a school is.

## Research Design

To investigate this hypothesis, I will conduct a cross-sectional study of the 2017 SEA results and assignment data. This data contains the students' primary school, 4 ranked choices of secondary schools, SEA scores and assigned secondary school and was provided to me by the Catholic Education Board.

### Data

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(readxl)
library(gt)
dat <- read_excel("data/data2017.xlsx")
gt(data.frame(head(dat)))
```

I will also add columns that identify the types of schools, whether religious boards, the government or private institutions manage the school. This data will be scraped from Trinidad and Tobago's Ministry of Education website that contains a list of all the schools in Trinidad and Tobago and their corresponding type [@noauthor_ministry_nodate]. I then recreated the assignment algorithm (See Appendix) that places primary school students into secondary schools to generate a new dataset where students are placed into secondary schools without the use of the 20% list.

In addition to this dataset, I will use another dataset from the Ministry of Education that contains cutoff scores for all secondary schools in 2017.

```{r}
cutoffs <- read_excel("data/cutoff_2017.xlsx")
gt(data.frame(head(cutoffs)))
```

The difference between the original assignments (with 20% list) and my generated assignments (without 20% list) with regards to the proportion of private primary school students in prestige secondary schools is displayed in Graph 1 below:

```{r, echo=FALSE, message=FALSE}
# data with new assignments
dat_n <- read_excel("data/masterdata_new_assignment.xlsx")

# prestige schools codes
prestige <- as.numeric(unique(dat_n$sec_no20list))

# useful columns
dat_n <- dat_n[, c("STUDENT PIN", "Gender", "Composite", "1st", "2nd", "3rd", "4th","Pri_sch_TYPE", "Gov_vs_denom", "sec_20list", "sec_no20list")]

# subset into prestige schools with 20% list assignments
dat_n <-  dat_n %>% mutate(list20_prestige = ifelse(sec_20list %in% prestige,1,0))
list20prestige_dat <- dat_n[dat_n$list20_prestige == 1, ]

# subset into prestige schools without 20% list assignments (0 for no list)
dat_n <-  dat_n %>% mutate(nolist20_prestige = ifelse(sec_no20list > 0 ,0,999))
nolist20prestige_dat <- dat_n[dat_n$nolist20_prestige == 0, ]
nolist20prestige_dat <- nolist20prestige_dat %>%
  select(-list20_prestige)

# rename list column
nolist20prestige_dat <- rename(nolist20prestige_dat, 
                               list20_prestige = nolist20_prestige)

# join into single dataset
cdat <- rbind(list20prestige_dat, nolist20prestige_dat)
# remove irrelevant assignments
cdat$sec_20list[cdat$list20_prestige == 0] <- NA
cdat$sec_no20list[cdat$list20_prestige == 1] <- NA


# pivot primary school type longer
cdat$x <- rep(1, length(cdat$Gender))
cdat <- cdat %>% pivot_wider(names_from = Pri_sch_TYPE,
                             values_from = x,
                             values_fill = 0)


ggplot(cdat) +
  geom_bar(aes(x=list20_prestige, y=mean(Private)), stat="identity", fill="tomato")+
  labs(x = "20% list",
       y = "Private school students",
       title = "Number of private school students in prestige schools")+
  scale_x_continuous(breaks=c(0,1))+
  theme_minimal()

```

*Graph 1 - This bar graph compares the number of students from private primary schools that are assigned to a prestige school with and without the 20% list in place. 1 represents the affirmative of the axis label, 0 represents the counterfactual.*

The graph demonstrates that there is a slightly greater number of private school students assigned to prestige schools with the 20% list in effect (20% list = 1) compared to the instance where the 20% list is not used (20% list = 0).

### Regression Models

Using private primary school attendance as a proxy for family wealth, I will estimate the strength of the relationship between the 20% list and rate of admittance of students from wealthy families to 'prestige' secondary schools.

I will estimate the effect with an Ordinary Least Squared (OLS) regression and a control for composite SEA scores (See Appendix for further discussion about controls).

$$ 20percentlist_{ij} =b_0 + b_1 (PrivPrestige_{ij}) + b_2(Composite_{ij}) +ε_i $$

$20percentlist_{ij}$: Binary variable describing whether a student's assignment to prestige school j came from the algorithm with the 20% list (1) or without the 20% list (0).

$b_0$: Intercept

$b_1$: The effect of the 20% list on the likelihood of a private school student attending a prestige school.

$PrivPrestige_{ij}$: Binary variable describing whether a student attended a private primary school and got accepted into a prestige school j (1) or not (0).

$b_2$: The effect of a student's composite SEA score on the likelihood of being admitted to a prestige school.

$Composite_{ij}$: Student's composite SEA score

$ε_{ij}$: Error term

## Findings

```{r, echo=FALSE, message=FALSE}
library(stargazer)
stargazer(lm(list20_prestige ~ Private + Composite,data = cdat),
          type = "text",
          title = "Regression Results",
          dep.var.labels = "20% list in effect")
```

## Empirical extensions

EXPLAIN!!! Heterogeneity

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# top 10 female
top10_f <- c(390,579,614,613,612,264,30,298,634,269)
top10_f_dat <- cdat[cdat$Gender == "F" & (cdat$sec_20list %in% top10_f | cdat$sec_no20list %in% top10_f), ]
a <- lm(list20_prestige ~ Private + Composite,data = top10_f_dat)

# top 10 male
top10_m <- c(389,456,260,457,276,27,634,213,470,598)
top10_m_dat <- cdat[cdat$Gender == "M" & (cdat$sec_20list %in% top10_m | cdat$sec_no20list %in% top10_m), ]
b <- lm(list20_prestige ~ Private + Composite,data = top10_m_dat)

# 11-20 female
top11to20_f <- c(265,63,64,276,504,598,138,141,31,419)
top11to20_f_dat <- cdat[cdat$Gender == "F" & (cdat$sec_20list %in% top11to20_f | cdat$sec_no20list %in% top11to20_f), ]
c <- lm(list20_prestige ~ Private + Composite,data = top11to20_f_dat)

# 11-20 male
top11to20_m <- c(138,622,504,668,141,692,584,669,362,67)
top11to20_m_dat <- cdat[cdat$Gender == "M" & (cdat$sec_20list %in% top11to20_m | cdat$sec_no20list %in% top11to20_m), ]
d <- lm(list20_prestige ~ Private + Composite,data = top11to20_m_dat)

# 21-30 female
top21to30_f <- c(362,678,540,595,469,230,12,164,67,102)
top21to30_f_dat <- cdat[cdat$Gender == "F" & (cdat$sec_20list %in% top21to30_f | cdat$sec_no20list %in% top21to30_f), ]
e <- lm(list20_prestige ~ Private + Composite,data = top21to30_f_dat)

# 21-30 male
top21to30_m <- c(263,548,678,230,164,578,102,706,12,690)
top21to30_m_dat <- cdat[cdat$Gender == "M" & (cdat$sec_20list %in% top21to30_m | cdat$sec_no20list %in% top21to30_m), ]
f <- lm(list20_prestige ~ Private + Composite,data = top21to30_m_dat)

cat(stargazer(a, b, c, d, e, f,
          type = "html",
          dep.var.labels = "Effect at different levels",
          column.labels = c("Top 10 F", "Top 10 M", "11-20 F", "11-20 M", "21-30 F", "21-30 M")))

```

Nothing significant. Likely too small of a sample. Schools take in anywhere from 0% - 20%. Looking for very small differences. Not surprised by findings

## Discussion and Conclusion

There are several limitations in this research design. For example, private school attendance is not the most effective proxy for wealth. Having more accurate wealth data such as family income would be useful for the internal validity of this study.

As a robustness check, I could test the strength of the relationship between the 20% list and the proportion of private school students who get into their first-choice schools. First-choice schools are similar to prestige schools in the sense that they are representative of a family's preferences but they are more subjective. If results are similar, an argument can be made that the relationship is stronger.
