---
title: "Project Proposal"
author: "Luke Gooding"

date: last-modified
institute: "University of Pennsylvania"
toc: true

link-citations: TRUE
bibliography: bib.bib

format:
  html:
    self-contained: TRUE
    code-fold: TRUE

editor: source
---

## Introduction

Trinidad and Tobago earned its independence in 1962. In that transition period from colony to independent state, education was used as a means of post-colonial nation building [@london_educational_1997]. The emerging education system, therefore, emphasized access and quality with the message being that education is central to the economic development of any nation [@london_educational_1997]. The country consequently has a robust and ever expanding public schooling system which, as of 2019, educates 92% of primary school students across 558 institutions and 88% of secondary school students across 199 institutions [@noauthor_education_2021].

Importantly, one of the major stakeholders in Trinidad and Tobago's education system are churches. Before independence, most of the nation's schools were run by various religious boards. Churches owned and managed these schools with autonomy and were allowed to charge enrolled students. With the goal of universal secondary education, the government took over the responsibility of funding these schools, which are now known as government 'assisted' or denominational schools [@london_educational_1997]. They standardized syllabi and exams across all the nation's public schools while allowing these religious boards to maintain administrative responsibility and their religious ethos. This arrangement is codified in The Concordat of 1960 [@noauthor_policy_nodate].

Having a publicly funded school system requires the government to assign graduating Primary School students to Secondary Schools. I refer to this process of assignment as the transition process. First, students choose 4 secondary schools they want to attend and rank them. Then, they do the Secondary Entrance Assessment (SEA). Based on their performance on the exam, they are placed into one of their 4 choices using an assignment algorithm (Appendix). If they perform poorly, they will be assigned to a school near to where they live.

However, according to the Concordat, denominational secondary schools have a special privilege:

*The Principals of Assisted Secondary Schools will make available a minimum of 80 per centum of the First Form entry places to those who, by passing the test, qualify on the results of the Common Entrance Examination for free secondary education...The Principals will be free to allocate up to 20 per centum, the remaining places as they see fit provided normally that the pass list of the Common Entrance Examination serves to provide the pupils. [@noauthor_policy_nodate]*

This part of the Concordat, colloquially known as the '20% list' is the subject of my research. Campbell @campbell_endless_1997 refers to the Concordat as one of many colonial holdovers present in our education system that perpetuates elitism. Coloquially, we refer to elite secondary schools as 'prestige schools'.

'Prestige schools' are secondary schools that are considered high performing and are usually ranked first in students' 4 choices. Some refer to these schools as 'first choice schools'. To operationalize this term, I have devised a definition:

*If the number of students that chose a school as their first choice exceeds the capacity of that school, this school will be classified as a 'prestige school'.*

I can then order schools based on cutoff scores (similar to the way the Assignment Algorithm works). For example, if 250 students choose school X but school X has a capacity of 100, then X will be considered a prestige school with a cutoff score equal to the score of the 100th best performing student. I believe my definition is suitable because I think 'prestige' as a construct is more about perception than reality. It is possible to develop a ranking system for secondary schools based on historical academic and extracurricular performance but for the purpose of my analysis, the aggregated perception of students and their parents is more meaningful, and this is proxied by their choices. These choices set up a system of demand and supply where some schools have a greater demand than others, creating 'prestige'. It is not my goal to explain the underlying reason for this 'prestige' but rather to identify it and investigate a particular externality.

These schools also tend to be government assisted schools with the 20% list privilege. It then begs the question how this 20% list is used today. Remember, these churches went from having control over 100% of their incoming students in a time when secondary education was expensive but in low demand to now only controlling 20% of their incoming students in a time when secondary education is free but in high demand.

Unfortunately, there are not many studies identifying and focusing on 'prestige schools' in particular. Jackson @jackson_can_2013, looked at a special subset of 'selective schools'. He showed that attending schools with higher-achieving peers increases academic performance and that short-run increases in peer achievement improve outcomes only at high-achieving selective schools.

A study conducted by the Inter-American Development Bank shows that there are no meaningful differences between the academic outcomes of government versus government assisted secondary schools between the ages of 10 and 15 in terms of dropout rates or examination performance at age 15 in Trinidad and Tobago [@beuermann_privately_2015]. This is despite the study also noting large differences in teacher and peer quality between school types. The research made no mention of 'prestige schools' which I theorize is an outlying subset of government assisted schools.

This work is intended to inform citizens and policy makers about the mechanisms behind the 20% list. If there is evidence of inequity, whether from this study or others, parents, teachers and policy makers should mobilize to address this issue. 

As Lisle @lisle_when_2009 stated: the future of policy on the use of secondary school entrance examinations lies not in further tinkering or experimentation without evidence, but in developing a robust ongoing monitoring and evaluation system to provide information on examination processes and outcomes.

More broadly, the persistence of the SEA exam and similar 11-plus examinations is the subject of much research:

*One of the best established findings of educational research is that students’ social class and their attainment at school are linked. This is a worldwide phenomenon. Students from more prestigious social class backgrounds or higher-income families tend to obtain higher marks and examination grades, and are then more likely to stay on in any education system that celebrates qualifications over equity.* [@doi:10.1177/2158244018825171]

Worldwide, 11-plus exams have been discredited from both a psychometric and philosophical point of view by education authorities committed to equality of opportunity and non-elitism [@mackenzie_elevenplus_1989]. MacKenzie @mackenzie_elevenplus_1989 posits that the system persists in Trinidad and Tobago as a necessary mechanism to resolve the incredible demand for prestige schools. It is only through the improvement of non-prestige schools and the change in attitude of citizens toward them can the conversation about removing SEA be possible. MacKenzie @mackenzie_elevenplus_1989  suggests that, in the short term, modifications to the SEA examination will go some way to mitigating its deleterious effects. That is what I hope to do with this research project.

### Theory and Hypothesis

The existence of 'prestige schools' has many effects on individuals navigating the transition process. Failure to get into a 'prestige school' can have negative emotional, social and psychological impact on students and hence the effort to avoid this failure would cause students to experience stress, anxiety and fear [@wreo31474]. Parents also often resort to investing in supplementary tutoring in preparation for the SEA exam [@barrow_supplementary_2012].

With such high stakes, the 20% list is a powerful tool. While its purpose is to allow principals to hand pick students who share the same religion and values as their institution, many have claimed that the 20% list is a tool to allow students of wealthy families to gain admittance to 'prestige schools'. I am interested in whether or not the 20% list benefits wealthier families and generates inequality.

I hypothesize that there is a significant relationship between the existence of the 20% list and rate of admittance of students from wealthy families to 'prestige' secondary schools. More specifically, I think that attending a private primary school increases your likelihood of being admitted to a prestige school.

Given that I do not have granular data of students' family income, I think private primary school attendance is a suitable proxy for family wealth because only 8% of students belong to families that are willing to incur the cost of private schools.

### Empirical Extension

I will also test the strength of the relationship in question at different levels modelled after Jackson's research [@jackson_can_2013]. He uses a different ordering system to rank schools by mean incoming SEA scores. He then makes 3 categories: 1 to 30, 30 through 90 and Above 90 and compares the effect at each level. While my ordering system uses cutoff scores, I can model his categorization to determine if there are different strengths of the relationship between the 20% list and the rate of private school student admissions at the 1 to 10, 11-20 and 21-30 levels.

Ranking schools based on cutoff scores is a useful index for competitiveness of a school for two reasons. Firstly, in order for a school to have a cutoff score, there must be more students who choose the school as their first choice than there are spots available. Secondly, higher cutoff scores means higher performing, more competitive students.

These regressions will aid in addressing heterogeneity in my research question. It is entirely plausible that attending a private primary school has varying effects depending on how competitive a school is. I will also separate the data into male and female students because cutoff scores are also gendered (see Table 2). This is so because schools are either all-male or all-female and those that are unisex have a set number of seats available for each gender.

## Research Design

To investigate this hypothesis, I will conduct a cross-sectional study of the 2017 SEA results and assignment data. This data contains the students' primary school, 4 ranked choices of secondary schools, SEA scores and assigned secondary school and was provided to me by the Catholic Education Board.

### Data

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(readxl)
library(gt)
dat <- read_excel("data/data2017.xlsx")
gt(data.frame(head(dat)))
```

*Table 1: Original Dataset from the Catholic Education Board*

I will also add columns that identify the types of schools, whether religious boards, the government or private institutions manage the school. This data will be scraped from Trinidad and Tobago's Ministry of Education website that contains a list of all the schools in Trinidad and Tobago and their corresponding type [@noauthor_ministry_nodate]. I then recreated the assignment algorithm (See Appendix) that places primary school students into secondary schools to generate a new dataset where students are placed into secondary schools without the use of the 20% list. This new dataset can be seen in Table 2 below:

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# data with new assignments
dat_n <- read_excel("data/masterdata_new_assignment.xlsx")

# prestige schools codes
prestige <- as.numeric(unique(dat_n$sec_no20list))

# useful columns
dat_n <- dat_n[, c("STUDENT PIN", "Gender", "Composite", "1st", "2nd", "3rd", "4th","Pri_sch_TYPE", "Gov_vs_denom", "sec_20list", "sec_no20list")]

# subset into prestige schools with 20% list assignments
dat_n <-  dat_n %>% mutate(list20_prestige = ifelse(sec_20list %in% prestige,1,0))
list20prestige_dat <- dat_n[dat_n$list20_prestige == 1, ]

# subset into prestige schools without 20% list assignments (0 for no list)
dat_n <-  dat_n %>% mutate(nolist20_prestige = ifelse(sec_no20list > 0 ,0,999))
nolist20prestige_dat <- dat_n[dat_n$nolist20_prestige == 0, ]
nolist20prestige_dat <- nolist20prestige_dat %>%
  select(-list20_prestige)

# rename list column
nolist20prestige_dat <- rename(nolist20prestige_dat, 
                               list20_prestige = nolist20_prestige)

# join into single dataset
cdat <- rbind(list20prestige_dat, nolist20prestige_dat)
# remove irrelevant assignments
cdat$sec_20list[cdat$list20_prestige == 0] <- NA
cdat$sec_no20list[cdat$list20_prestige == 1] <- NA


# pivot primary school type longer
cdat$x <- rep(1, length(cdat$Gender))
cdat <- cdat %>% pivot_wider(names_from = Pri_sch_TYPE,
                             values_from = x,
                             values_fill = 0)

gt(cdat[sample(nrow(cdat), 10), ])
```
*Table 2: 10 random rows from dataset*

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(vtable)
st(cdat, vars = c('Private','sec_20list', 'sec_no20list', 'Composite'))
```
*Table 3: Summary table of variables of interest*

In addition to this dataset, I will use another dataset from the Ministry of Education that contains cutoff scores for all secondary schools in 2017.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
cutoffs <- read_excel("data/cutoff_2017.xlsx")
gt(data.frame(head(cutoffs)))
```
*Table 4: Cutoff Scores 2017*

### Regression Models

Using private primary school attendance as a proxy for family wealth, I will estimate the strength of the relationship between the 20% list and rate of admittance of students from wealthy families to 'prestige' secondary schools. I chose to include composite score as a control because it acts as a confounder in the sense that a higher composite grade increases the chance of being admitted to a prestige school. In order to isolate the effect of attending a private primary school, I intend to control for any factors that could be considered when selecting students for the 20% list such as a student's composite score. While I did not identify any other variables from my data as cofounders (See Appendix), it is entirely possible that unobserved ones exist. For example, it is plausible that family connections and social network can influence can act as a confounder. Relationships with primary both primary and secondary school faculty can play an important role in which secondary school a student is assigned to. Also, family religion is an important factor in determining which primary school a student attends and it plays a role in the 20% list decision-making process.


Therefore, the model I choose to investigate my hypothesis is an Ordinary Least Squared (OLS) regression with a control for composite SEA scores (See Appendix for further discussion about controls).

$$ 20percentlist_{ij} =b_0 + b_1 (PrivPrestige_{ij}) + b_2(Composite_{ij}) +ε_i $$

$20percentlist_{ij}$: Binary variable describing whether a student's assignment to prestige school j came from the algorithm with the 20% list (1) or without the 20% list (0).

$b_0$: Intercept

$b_1$: The effect of the 20% list on the likelihood of a private school student attending a prestige school.

$PrivPrestige_{ij}$: Binary variable describing whether a student attended a private primary school and got accepted into a prestige school j (1) or not (0).

$b_2$: The effect of a student's composite SEA score on the likelihood of being admitted to a prestige school.

$Composite_{ij}$: Student's composite SEA score

$ε_{ij}$: Error term

## Findings

The difference between the original assignments (with 20% list) and my generated assignments (without 20% list) with regards to the proportion of private primary school students in prestige secondary schools is displayed in Graph 1 below:

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(cdat) +
  geom_bar(aes(x=list20_prestige, y=mean(Private)), stat="identity", fill="tomato")+
  labs(x = "20% list",
       y = "Private school students",
       title = "Number of private school students in prestige schools")+
  scale_x_continuous(breaks=c(0,1))+
  theme_minimal()

```

*Graph 1 - This bar graph compares the number of students from private primary schools that are assigned to a prestige school with and without the 20% list in place. 1 represents the affirmative of the axis label, 0 represents the counterfactual.*

The graph demonstrates that there is a slightly greater number of private school students assigned to prestige schools with the 20% list in effect (20% list = 1) compared to the instance where the 20% list is not used (20% list = 0). This gives merit to my hypothesis as it demonstrates an observable difference between the worlds where the 20% list exists and does not exist.

Using the regression models outlined in the previous section, I will determine whether this difference is significant.

###Primary Regression

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
library(stargazer)
stargazer(lm(list20_prestige ~ Private + Composite,data = cdat),
          type = "html",
          title = "Regression Results",
          dep.var.labels = "20% list in effect")
```

According to this regression, when the 20% list is in effect, private school students have a 2.6% greater chance than non-private school students of being assigned to a prestige school. This effect is NOT significant at the p<0.05 level when students' composite SEA scores are controlled for. Therefore, I am not able to find a statistically significant relationship and fail to reject the null to my hypothesis. The evidence does not suggest that attending a private primary school increases the likelihood of being assigned to a prestige school.

### Empirical Extension Results

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
# top 10 female
top10_f <- c(390,579,614,613,612,264,30,298,634,269)
top10_f_dat <- cdat[cdat$Gender == "F" & (cdat$sec_20list %in% top10_f | cdat$sec_no20list %in% top10_f), ]
a <- lm(list20_prestige ~ Private + Composite,data = top10_f_dat)

# top 10 male
top10_m <- c(389,456,260,457,276,27,634,213,470,598)
top10_m_dat <- cdat[cdat$Gender == "M" & (cdat$sec_20list %in% top10_m | cdat$sec_no20list %in% top10_m), ]
b <- lm(list20_prestige ~ Private + Composite,data = top10_m_dat)

# 11-20 female
top11to20_f <- c(265,63,64,276,504,598,138,141,31,419)
top11to20_f_dat <- cdat[cdat$Gender == "F" & (cdat$sec_20list %in% top11to20_f | cdat$sec_no20list %in% top11to20_f), ]
c <- lm(list20_prestige ~ Private + Composite,data = top11to20_f_dat)

# 11-20 male
top11to20_m <- c(138,622,504,668,141,692,584,669,362,67)
top11to20_m_dat <- cdat[cdat$Gender == "M" & (cdat$sec_20list %in% top11to20_m | cdat$sec_no20list %in% top11to20_m), ]
d <- lm(list20_prestige ~ Private + Composite,data = top11to20_m_dat)

# 21-30 female
top21to30_f <- c(362,678,540,595,469,230,12,164,67,102)
top21to30_f_dat <- cdat[cdat$Gender == "F" & (cdat$sec_20list %in% top21to30_f | cdat$sec_no20list %in% top21to30_f), ]
e <- lm(list20_prestige ~ Private + Composite,data = top21to30_f_dat)

# 21-30 male
top21to30_m <- c(263,548,678,230,164,578,102,706,12,690)
top21to30_m_dat <- cdat[cdat$Gender == "M" & (cdat$sec_20list %in% top21to30_m | cdat$sec_no20list %in% top21to30_m), ]
f <- lm(list20_prestige ~ Private + Composite,data = top21to30_m_dat)

stargazer(a, b, c, d, e, f,
          type = "html",
          dep.var.labels = "Effect at different levels",
          column.labels = c("Top 10 F", "Top 10 M", "11-20 F", "11-20 M", "21-30 F", "21-30 M"))

```

Similar to the primary regression, these regression results show that when the 20% list in in effect, private primary school students are not more likely to attend a prestige school. This table in particular demonstrates that this lack of an effect holds for 3 different levels of prestige secondary schools and across sex.

There are several factors that may explain why these regressions could not explain the difference noted in Graph 1. Firstly, it may be the case that prestige schools do not demonstrate preference for private school students. However, due to the fact that an effect was noted at the p\<0.1 level, it is possible that an effect exists but my study design does not have enough power due to a small sample size. Recall that prestige schools can choose 0%-20% of their incoming class and only 8% of students taking the SEA exam are private school students. This suggests that the effect I am looking for is likely incredibly small and may only be observable with a larger dataset (for example, results and assignments from other years aside from 2017).

## Discussion and Conclusion

This study has not yielded significant results. As explained above, it could be the case that my hypothesis is not based in reality or it could be the case that the study design is faulty and requires greater statistical power. Additionally, private school attendance is not the most effective proxy for wealth. Having more accurate wealth data such as family income would be useful for the internal and external validity of this study. Also, I am only using 2017 SEA data. Therefore, I am concerned about the external validity of my findings, particularly because the Concordat has been around since the 1960s. It is possible that my findings are unique to 2017 and is not representative of other years.

Future research should use a larger dataset and better measures of family wealth. Also, a qualitative component that gets at the heart of the 20% list decision making process could prove invaluable such as a thematic analysis of interviews with past and present principals who have created 20% lists for their institutions.

It is my hope that I can add to the conversation about the Concordat and the 20% list in a meaningful, empirical way. If it turns out that the Concordat poses a threat to equity or perpetuates a system of elitism, I hope to be a voice of change. At the least, I hope that this study can demystify a currently opaque process.

